Quasi-universal acceptance of basic science in the US

### 2025-07-07 paper.Rmd
second round revisions

  Many people report a low degree of trust in science, or endorse conspiracy theories that violate basic scientific knowledge. This might indicate a wholesale rejection of science in substantial segments of the population. In four studies, we asked 782 US participants questions about trust in science, conspiracy beliefs, and basic science (e.g. the relative size of electrons and atoms). Participants were provided with the scientifically consensual answer to the basic science questions, and asked whether they accept it. Acceptance of the scientific consensus was very high in the sample as a whole (95.1%), but also in every sub-sample (e.g., no trust in science: 87.3%; complete endorsement of flat Earth theory: 87.2%). This quasi-universal acceptance of basic science suggests that people are motivated to reject specific scientific beliefs, and not science as a whole. This could be leveraged in science communication.
keywords          : Trust in science; conspiracy theories; conspiracy thinking; science knowledge.
floatsintext      : yes
linenumbers       : no 
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "doc" # "doc" for nice look, "man" for manuscripty
output: 
  papaja::apa6_pdf: 
    keep_tex: true
  papaja::apa6_docx: default
always_allow_html: true
appendix:
  - "appendix_acceptance_knowledge.Rmd"
  - "appendix_conspiracy_thinking.Rmd"
  - "appendix_trust.Rmd"
  - "appendix_conspiracy_belief.Rmd"
  - "appendix_knowledge.Rmd"
  - "appendix_by-question.Rmd"
  - "appendix_conspiracy-scales.Rmd"
  - "appendix_exp1.Rmd"
  - "appendix_exp2.Rmd"
  - "appendix_exp3.Rmd"
  - "appendix_exp4.Rmd"
bibliography: references.bib
---

@@

Although recent global evidence shows that trust in science is moderately high [@colognaTrustScientistsTheir2024], it is far from being at ceiling. Large-scale polls have shown that people who report a high degree of trust in science are a minority in most countries, and they are outnumbered by people who have low trust in science in many areas, e.g., most of Africa and significant parts of Asia [@wellcomeglobalmonitorWellcomeGlobalMonitor2018; @wellcomeglobalmonitorWellcomeGlobalMonitor2020]. Moreover, trust in science has recently been declining in some countries [@alganTrustScientistsTimes2021; @brianAmericansTrustScientists2023; although see @wellcomeglobalmonitorPublicTrustScientists2021; and @funkPublicConfidenceScientists2020] and in the US it is increasingly polarizing [@gauchatPoliticizationSciencePublic2012; @krauseTrendsAmericansTrust2019a; @liPolarizationPublicTrust2022].

Besides low answers on general trust in science questions, another indicator of distrust in science is the belief in conspiracy theories that question the scientific consensus on issues such as vaccination, climate change, and even the shape of the Earth. Conspiracy theories–in the realm of science or elsewhere–typically accuse a small group of powerful people to pursue nefarious goals in secrecy [@douglasUnderstandingConspiracyTheories2019; @medeSciencerelatedPopulismConceptualizing2020]. Some of these conspiracy theories are widespread [@rutjensConspiracyBeliefsScience2022]. In 2023, a survey in eight different countries found that up to 24% of respondents agreed that “climate change is a hoax and scientists touting its existence are lying” [@stockemerUnderstandingClimateChange2024]. Similarly, in 2021, 40% of Americans believed that “the dangers of genetically-modified foods are being hidden from the public” [down from 45% in 2020, @uscinskiHaveBeliefsConspiracy2022]. People who hold such views not only reject the relevant scientifically consensual facts, but also tend to believe in other conspiracy theories [@lewandowskyRoleConspiracistIdeation2013; @hornseyPsychologicalRootsAntivaccination2018; @hornseyRelationshipsConspiratorialBeliefs2018], and tend to say that they distrust science more generally [@vranicDidMyOwn2022; @stockemerUnderstandingClimateChange2024]. In spite of these correlations, the causal relationship between declaring general distrust in science and believing in one or several anti-science conspiracy theories is not clear. Although conspiracy thinking has been identified as a "root cause" of anti-science attitudes [@hornseyAttitudeRootsJiu2017] in the past, these claims rest largely on observational data.

@@

Studies 1 and 2 were conducted on the standard sample of US participants recruited on the platform Prolific Academic. In order to increase the share of participants with low trust in science, and who endorse conspiracy theories, Studies 3 and 4 used the same platform, but only recruited participants who had declared previously being skeptical of vaccination.

@@

In all analyses reported in the main paper, we measure trust in science via a question selected from the Wellcome Global Monitor surveys [@wellcomeglobalmonitorWellcomeGlobalMonitor2018; @wellcomeglobalmonitorWellcomeGlobalMonitor2020]: "In general, would you say that you trust science a lot, some, not much, or not at all? [1 = Not at all, 2 = Not much, 3 = Some, 4 = A lot]". We chose this question as it seemed to be the most general one. In the ESM (sections \@ref(exp1) to \@ref(exp4)), we additionally report results for two alternative measures of trust included in our studies: Another from the WGM surveys ("How much do you trust scientists in this country? Do you trust them a lot, some, not much, or not at all? [1 = Not at all, 2 = Not much, 3 = Some, 4 = A lot]"), and one from the Pew Research Center [e.g. @funkKeyFindingsPublic2019] ("How much confidence do you have in scientists to act in the best interests of the public? [1-5; 1 = No confidence at all, 5 = A great deal of confidence]"), the latter having been used in a recent international study on trust in science [@colognaTrustScientistsTheir2024]. We selected these items so that we could compare the answers in our sample to global survey results. We find that all three items are highly correlated throughout all studies, and that our results reported here generally replicate when using either of the alternatives measures[^2] (see ESM).

@@

n an exploratory analysis, we find that

While actual acceptance of scientific knowledge is only weakly correlated with trust in science,

@@

In applied terms, the present results have implications for science communication. Across various domains of science knowledge such as climate change [@veckalov27countryTestCommunicating2024] or vaccination [@salmonVaccineHesitancyCauses2015], researchers have observed a consensus gap: a gap between the scientific consensus and public opinion. Some of the most worrying consensus gaps relate to climate change, as substantial segments of the population disagree with scientists on what is happening and what to do about it [e.g., @eganClimateChangeUS2017]. Yet, we found that for much of basic, non-contentious science, such consensus gaps are absent: The vast majority of our participants--including two vaccine-skeptical samples (Studies 3 and 4)--did not appear to have general grounds for distrusting science, which should have led them to reject most or all of the science knowledge presented to them. Since even people who say they distrust science, or who reject specific scientific facts appear to accept most of basic science, stressing the basic science components of publicly controversial fields, from GMOs to climate change, might help reduce the consensus gaps observed in these domains [on climate change, see, e.g. @ranneyClimateChangeConceptual2016].

The present studies have a number of limitations, in particular the lack of representative samples, and the focus on a single country. While our results support motivated reasoning accounts of trust in science, they leave important questions unaddressed, in particular: Why do people reject specific scientific findings but not others? Suggestions have already been made for a number of issues such as vaccination [@mitonCognitiveObstaclesProVaccination2015], GMOs [@blanckeFatalAttractionIntuitive2015], or nuclear energy [@hacquinDisgustSensitivityPublic2021]. However, research is still needed to better understand what motivates these rejections [see e.g., @hornseyWhyFactsAre2020]. Moreover, rejection of scientific facts can manifest in various ways, for example calling into question the integrity of scientists [@medeSciencerelatedPopulismConceptualization2023a; @medeWhoSupportsSciencerelated2022], or denying the very possibility of scientifically investigating certain issues [@munroScientificImpotenceExcuse2010]. The classification of justifications we present here does not address these processes in detail. Finally, by design, most of the basic science knowledge presented in this study didn't directly relate to anything controversial for most people. Future studies could look at basic science that does relate to scientific topics which are the object of public and policy-relevant debates (e.g. GM food, climate science, vaccines), or scientific findings for which there is a more obvious potential conflict of interest (e.g. non-publicly funded medical research).

>>>>

  Substantial minorities of the population report a low degree of trust in science, or endorse conspiracy theories that violate basic scientific knowledge. This might indicate a wholesale rejection of science. In four studies, we asked 782 US participants questions about trust in science, conspiracy beliefs, and basic science (e.g. the relative size of electrons and atoms). Participants were provided with the scientifically consensual answer to the basic science questions, and asked whether they accept it. Acceptance of the scientific consensus was very high in the sample as a whole (95.1%), but also in every sub-sample (e.g., no trust in science: 87.3%; complete endorsement of flat Earth theory: 87.2%). This quasi-universal acceptance of basic science suggests that people are motivated to reject specific scientific beliefs, and not science as a whole. This could be leveraged in science communication.
keywords          : Trust in science; conspiracy theories; conspiracy thinking; science knowledge.
floatsintext      : yes
linenumbers       : no 
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "doc" # "doc" for nice look, "man" for manuscripty
output: 
  papaja::apa6_docx: default
  papaja::apa6_pdf: 
    keep_tex: true
always_allow_html: true
appendix:
  - "appendix_acceptance_knowledge.Rmd"
  - "appendix_conspiracy_thinking.Rmd"
  - "appendix_trust.Rmd"
  - "appendix_conspiracy_belief.Rmd"
  - "appendix_knowledge.Rmd"
  - "appendix_by-question.Rmd"
  - "appendix_conspiracy-scales.Rmd"
  - "appendix_exp1.Rmd"
  - "appendix_exp2.Rmd"
  - "appendix_exp3.Rmd"
  - "appendix_exp4.Rmd"
bibliography: references.bib
---

@@

Although recent global evidence shows that trust in science is moderately high [@colognaTrustScientistsTheir2025], it is far from being at ceiling. Large-scale polls have shown that people who report a high degree of trust in science are a minority in most countries, and they are outnumbered by people who have low trust in science in many areas, e.g., most of Africa and significant parts of Asia [@wellcomeglobalmonitorWellcomeGlobalMonitor2018; @wellcomeglobalmonitorWellcomeGlobalMonitor2020]. Moreover, trust in science has recently been declining in some countries [@alganTrustScientistsTimes2021; @brianAmericansTrustScientists2023; although see @wellcomeglobalmonitorPublicTrustScientists2021; and @funkPublicConfidenceScientists2020] and in the US it is increasingly polarizing [@gauchatPoliticizationSciencePublic2012; @krauseTrendsAmericansTrust2019a; @liPolarizationPublicTrust2022].

Besides low answers on general trust in science questions, another indicator of distrust in science is the belief in conspiracy theories that question the scientific consensus on issues such as vaccination, climate change, and even the shape of the Earth. Conspiracy theories–in the realm of science or elsewhere–typically accuse a small group of powerful people to pursue nefarious goals in secrecy [@douglasUnderstandingConspiracyTheories2019; @medeSciencerelatedPopulismConceptualizing2020]. Some of these conspiracy theories are widespread [@rutjensConspiracyBeliefsScience2022]. Data from the European Social Survey between 2020 and 2022 suggests that in several European countries, relatively large minorities of the population believed that COVID-19 was a result of deliberate and concealed efforts of some government or organisations: around 30% in Spain and Portugal, 22% in the UK, and 25% in Poland [@jabkowskiExploringCOVID19Conspiracy2023]. In 2023, a survey in eight different countries found that up to 24% of respondents agreed that “climate change is a hoax and scientists touting its existence are lying” [@stockemerUnderstandingClimateChange2024]. Similarly, in 2021, 40% of Americans believed that “the dangers of genetically-modified foods are being hidden from the public” [down from 45% in 2020, @uscinskiHaveBeliefsConspiracy2022]. People who hold such views not only reject the relevant scientifically consensual facts, but also tend to believe in other conspiracy theories [@lewandowskyRoleConspiracistIdeation2013; @hornseyPsychologicalRootsAntivaccination2018; @hornseyRelationshipsConspiratorialBeliefs2018; @littrellBeliefScienceRelatedConspiracy2025], and tend to say that they distrust science more generally [@vranicDidMyOwn2022; @stockemerUnderstandingClimateChange2024]. In spite of these correlations, the causal relationship between declaring general distrust in science and believing in one or several anti-science conspiracy theories is not clear. Although conspiracy thinking has been identified as a "root cause" of anti-science attitudes [@hornseyAttitudeRootsJiu2017] in the past, these claims rest largely on observational data.

@@

The surveys were hosted via Qualtrics and participants were recruited via Prolific Academic, a crowdsourcing platform that allows researchers to pre-screen and invite eligible participants from their participant pool. Studies 1 and 2 were conducted on the standard sample of US participants. In order to increase the share of participants with low trust in science, and who endorse conspiracy theories, Studies 3 and 4 used the same platform, but only recruited participants who had declared previously being skeptical of vaccination.

Because recruitment was conducted through Prolific, we do not have access to the number of individuals who viewed the study invitation but chose not to participate. As such, it is not possible for us to calculate response rates (the proportion of invited participants who completed the survey) for the studies. However, we employed attention checks and, for each study, we report exclusions based on these checks. Moreover, the study invitation was purposefully nondescript (“short psychology experiment”), so that it would be impossible for participants to self-select because of their attitudes towards science for instance.

@@

In all analyses reported in the main paper, we measure trust in science via a question selected from the Wellcome Global Monitor surveys [@wellcomeglobalmonitorWellcomeGlobalMonitor2018; @wellcomeglobalmonitorWellcomeGlobalMonitor2020]: "In general, would you say that you trust science a lot, some, not much, or not at all? [1 = Not at all, 2 = Not much, 3 = Some, 4 = A lot]". We chose this question as it seemed to be the most general one. In the ESM (sections \@ref(exp1) to \@ref(exp4)), we additionally report results for two alternative measures of trust included in our studies: Another from the WGM surveys ("How much do you trust scientists in this country? Do you trust them a lot, some, not much, or not at all? [1 = Not at all, 2 = Not much, 3 = Some, 4 = A lot]"), and one from the Pew Research Center [e.g. @funkKeyFindingsPublic2019] ("How much confidence do you have in scientists to act in the best interests of the public? [1-5; 1 = No confidence at all, 5 = A great deal of confidence]"), the latter having been used in a recent international study on trust in science [@colognaTrustScientistsTheir2025]. We selected these items so that we could compare the answers in our sample to global survey results. We find that all three items are highly correlated throughout all studies, and that our results reported here generally replicate when using either of the alternatives measures[^2] (see ESM).

@@

@@

In applied terms, the present results have implications for science communication. Across various domains of science knowledge such as climate change [@veckalov27countryTestCommunicating2024] or vaccination [@salmonVaccineHesitancyCauses2015], researchers have observed a consensus gap: a gap between the scientific consensus and public opinion. Some of the most worrying consensus gaps relate to climate change, as substantial segments of the population disagree with scientists on what is happening and what to do about it [e.g., @eganClimateChangeUS2017]. Yet, we found that for much of basic, non-contentious science, such consensus gaps are absent: The vast majority of our participants--including two vaccine-skeptical samples (Studies 3 and 4)--did not appear to have general grounds for distrusting science, which should have led them to reject most or all of the science knowledge presented to them. Since even people who say they distrust science, or who reject specific scientific facts appear to accept most of basic science, stressing the basic science components of publicly controversial fields, from GMOs to climate change, might help reduce the consensus gaps observed in these domains [on climate change, see, e.g. @ranneyClimateChangeConceptual2016]. Past research has shown that arguments building on specific scientific findings are more convincing even to populists who are generally skeptical towards expertise [@peresmanPopulistsListenExpertise2025].

Why do most people (in the US) trust most of basic science? Based on the present studies, we can only speculate. But if this trust has a rational basis, then education–and science education in particular–is the most likely explanation. Large-scale surveys have identified science education as one of the strongest, or the strongest, correlates of trust in science [@noyScienceGoodEffects2019; @wellcomeglobalmonitorWellcomeGlobalMonitor2018] If these studies were to assess trust in basic scientific knowledge, rather than assessing trust in science more broadly, we might expect effects to be even stronger. Differentiating the impact of education on different facets of trust would be an interesting avenue for future research.

The present studies have a number of limitations, in particular the lack of representative samples, and the focus on a single country. The use of online samples recruited via Prolific does not allow us to calculate response rates and thus estimate non-response bias. However, participants could not have self-selected on the basis of the study topic and, by targeting vaccine-skeptic samples in some of the studies, we made an effort in recruiting those participants that can be assumed to be most reluctant to partake in a study confronting them with science facts. While our results support motivated reasoning accounts of trust in science, they leave important questions unaddressed, in particular: Why do people reject specific scientific findings but not others? Suggestions have already been made for a number of issues such as vaccination [@mitonCognitiveObstaclesProVaccination2015], GMOs [@blanckeFatalAttractionIntuitive2015], or nuclear energy [@hacquinDisgustSensitivityPublic2021]. However, research is still needed to better understand what motivates these rejections [see e.g., @hornseyWhyFactsAre2020]. Moreover, rejection of scientific facts can manifest in various ways, for example calling into question the integrity of scientists [@medeSciencerelatedPopulismConceptualization2023a; @medeWhoSupportsSciencerelated2022], or denying the very possibility of scientifically investigating certain issues [@munroScientificImpotenceExcuse2010]. The classification of justifications we present here does not address these processes in detail. Finally, by design, most of the basic science knowledge presented in this study didn't directly relate to anything controversial for most people. Future studies could look at basic science that does relate to scientific topics which are the object of public and policy-relevant debates (e.g. GM food, climate science, vaccines), or scientific findings for which there is a more obvious potential conflict of interest (e.g. non-publicly funded medical research).
